\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing

### 2012-10-12 ace/ace.tex
Renamed ipdp13 to ace

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[10pt, conference, compsocconf]{IEEEtran}
% Add the compsocconf option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

\usepackage{cite}
% IEEE wants to use ],\,[ for this
% but that looks dumb...
\renewcommand{\citepunct}{,\,}

%\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

@@

\usepackage{color}

@@

\author{\IEEEauthorblockN{Cyrus Omar, Nathan Fulton, Sharschchandra Bidargaddi and Jonathan Aldrich}
\IEEEauthorblockA{School of Computer Science\\
Carnegie Mellon University\\
Pittsburgh, PA, USA\\
\url{http://www.acelang.org/}
}}

@@

\begin{abstract}
We introduce the Ace compilation environment and demonstrate its practicality as a foundational tool for both high-performance computing research and in practice. 
Ace consists of a statically-typed programming language with user-extensible semantics controlled by a compile-time meta\-language, Python. Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}. This mechanism allowed us to express the primitives of the C99 and OpenCL languages as simple modules within Ace. We then give examples of several first-class parallel abstractions that can be defined and used alongside these basic primitives, demonstrating the foundational role that an actively-typed system like Ace can play for researchers and domain experts developing first-class language abstractions, as well as practitioners wishing to use multiple such abstractions within a single program. Active library support in Ace also supports more general forms of type-aware metaprogramming, and Ace functions can be compiled and launched directly from Python, taking its standard numeric data structures as arguments.
Using active library support in Ace, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace can be used, at scale, in practice today.
\end{abstract}

%\begin{IEEEkeywords}
%programming languages, heterogeneous computing, metaprogramming, code generation
%\end{IEEEkeywords}

@@

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware and about 20\% of scientists target either local clusters or super\-computers for more numerically-intensive computations as well \cite{hannay2009scientists}. To fully harness the power of this modern hardware, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-intensive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For tasks where these interpretted languages are too slow, they generally call into code written in a statically-typed, low-level language like C and Fortran and use low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers. {\color{red} CITATION?}

The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns relating to performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users have not commonly adopted these approaches, and many have become generally skeptical of new approaches originating in the research community. This viewpoint was perhaps most succinctly expressed by a participant interviewed in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of scientific and high-performance computing.

One reason for this gap between research and practice is because new features and abstractions are often developed and distributed together with a new programming language. Some prominent examples include:
\begin{itemize}
\item UPC, Fortress, X10 and Chapel, each featuring some form of first-class partitioned global address space (PGAS) support;
\item Erlang, Charm++, Occam and Axum, featuring some form of first-class support for message-passing; 
\item Cilk and NESL, featuring nested data parallelism
\item {\color{red} should I flesh this out more? less?}
\end{itemize}

Unfortunately, using a domain-specific research language comes with a host of challenges. {\color{red} List challenges here?} Indeed, the most widely adopted parallel programming abstractions have been those that are distributed as libraries, such as MapReduce, or as simple, conservative language extensions available in existing compilers, such as OpenMP.

Ace is a language-integrated compilation environment that aims to make it easier for researchers to introduce new first-class language constructs as libraries. A {\em compilation environment} is a framework that allows users to programmatically control the compilation of programs. In Ace, users can interleave two kinds of program logic: compile-time and run-time. The compile-time logic, written in the {\em metalanguage}, is the top-level code in each file. Here, users can define and import new primitive types and operators, generate functions programmatically, perform arbitrary code transformations and emit functions for later execution. Ace also permits an interactive mode of use whereby the functions can also be executed immediately after being defined, a common usage scenario for professional end-users.

Ace is an {\it actively-typed} compilation environment. This means that types are first-class objects in the metalanguage, and that the logic needed to typecheck and compile the operators associated with a type is defined alongside the type itself. The process by which this occurs is called {\em active typechecking and translation (\ATT)}, and we discuss it in detail in Section \ref{att}. \ATT~is flexible enough that we are able to define the entirety of the C99 and OpenCL languages   with it, demonstrating its practicality and providing a foundation for higher-level abstractions that we discuss in Section \ref{abstractions}. Notably, these abstractions can be imported piecemeal as libraries -- they are not defined as new languages, nor as extensions for particular compilers.

Although the \ATT~mechanism is designed for use by researchers and domain experts, Ace is designed to be practical for professional end-user developers as well. It is based on the Python programming language, which is among the most broadly adopted languages in these communities. Indeed, the language itself is implemented entirely as a Python library -- no new executables need to be installed. Extension writers can target arbitrary backend languages and use existing compilers for back-end optimization. In this paper, we demonstrate the use of the flexible OpenCL platform for this purpose. Using these capabilities, we developed a

In addition to the general introduction of Ace as a foundational tool for research and practice and generalizable lessons relating to language design that we note below, we highlight several individual contributions in this paper:

\begin{itemize}
\item \ATT, a generalization of active libraries, where types are metalanguage objects and programs are given semantics by type-associated methods called by the compiler.
\item A method for eliminating type annotations by using a form of type inference within functions and type propagation between functions.
\item A method to check the correctness of generated code, by checking for satisfaction of representational consistency constraints associated with types, defined in Section \ref{repcon}. 
\item Several type-aware code generation and simulation orchestration techniques used in the clegans library, described in Section \ref{clegans}.
\end{itemize}

@@

%We then introduce a new programming language named $\cloquence$ (pronounced ``C eloquence'') in Section 3. $\cloquence$ initially targets code paths where low-level languages, particularly OpenCL, would be used today. By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that $\cloquence$ may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
%
%Unlike many existing languages, $\cloquence$ is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}. We argue that this feature may make $\cloquence$ particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
%
%In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem. Finally, conclude in Section 6 with future directions for both the $\cloquence$ language and the research community.
%

\section{ACE: An \underline{A}ctively-Typed \underline{C}ompilation \underline{E}nvironment}
Ace is based on the Python programming language and implemented entirely as a Python package. The metalanguage is Python itself, the object language uses Python's grammar directly, and the compiler is implemented in Python as well. Write some more here. Why Python? Popularity and to aid in adoption: can be installed just like any other Python module, uses Python's module system and distribution infrastructure, piggybacks on existing Python toolchain.

We now describe a programming language designed with many of the criteria described in the previous sections in mind, $\cloquence$. $\cloquence$ is built around a minimal core and nearly every construct in the language is implemented as a library. Rather than beginning directly with this extension mechanism, however, let us begin with a concrete example of the language in use.

\subsection{Example: Higher-Order Map for OpenCL}
Figure \ref{map} shows the standard data-parallel map function written using an extension that implements the full OpenCL language as a library. Each thread, indexed by \verb|gid|, applies the transfer function, \verb|fn|, to the  corresponding element of the input array, \verb|in|, writing the result into the corresponding location in the output array, \verb|out|.

\subsection{Syntax}
Readers familiar with the Python programming language will recognize the style of syntax used in Figure \ref{map}. In fact, $\cloquence$ uses the Python grammar and parsing facilities directly. Several factors motivated this design decision. First, Python's syntax is widely credited as being particularly simple and readable, due to its use of significant whitespace and conventional mathematical notation. Python is one of the most widely-used languages in scientific computing, so its syntax is already familiar to much of the field. And significantly, a large ecosystem of tools already exist that work with Python files, such as code editors, syntax highlighters, style checkers and documentation generators. These can be used without modification to work with $\cloquence$ files. Therefore, by re-using an existing, widely-used grammar, we are able to satisfy many of the design criteria described in Section \ref{syntax} and the adoption criteria described in Section \ref{tools} without significant development effort.

\begin{figure}
\small{
\begin{verbatim}
from ace.OpenCL import OpenCL, get_global_id

@OpenCL.fn
def map(in, out, fn):
    gid = get_global_id(0)
    out[gid] = fn(in[gid])
\end{verbatim}
}
\vspace{-10pt}
\caption{A generic data-parallel map function written using the user-defined Ace OpenCL module.}
\label{map}
\end{figure}
\begin{figure}
\small{
\begin{verbatim}
from figure1 import map
from ace.OpenCL import global_ptr, double, sin

T = global_ptr(double)
map_sin_dbl = map.compile(T, T, sin.cl_type)
\end{verbatim}
}
\vspace{-10pt}
\caption{The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers. The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
\label{mapsindblace}
\end{figure}
\begin{figure}
\small{\begin{verbatim}
#pragma OPENCL EXTENSION cl_khr_fp64 : enable

__kernel void map_sin_dbl(
  __global double* in, 
  __global double* out) 
{
    size_t gid;
    gid = get_global_id(0);
    out[gid] = sin(in[gid]);
}\end{verbatim}
}
\vspace{-10pt}
\caption{The OpenCL kernel function generated by running \texttt{acecc figure2.py}, or by executing the  \texttt{print map\_sin\_dbl.code} statement.}
\label{mapsinint}
\vspace{-10pt}
\end{figure}

\subsection{Semantics}
$\cloquence$ is a compiled, statically-typed language. The extension used in Figure \ref{map} implements the full OpenCL type system, which includes features inherited from C99 like pointer arithmetic and OpenCL-specific constructs like vector types. In fact, compilation of Figure \ref{map} will produce code semantically equivalent, and with the same performance profile, as manually-written OpenCL code. Compilation produces OpenCL source code, so $\cloquence$, when used with this extension, inherits OpenCL's portability profile as well, as described in Section \ref{portability}.

In OpenCL itself, however, there is no way to write a function like \verb|map|. As described in Section \ref{hof}, OpenCL lacks support for higher-order functions or polymorphism (over the types of \verb|in| and \verb|out| in this case.) To map a function over an array, developers must create a separate function for every combination of argument types and for every transfer function. Figure 2 shows an OpenCL kernel that implements a map specifically for the \verb|sin| function applied to arrays of doubles in global memory. Note that because OpenCL does not support templates or function pointers, this is the most general solution possible without explicit code generation. However, even in CUDA, which does support templates, there is greater syntactic overhead for specializing a function with particular types than in $\cloquence$, where it happens implicitly.

To support this much more concise form for specifying kernels, $\cloquence$ uses techniques pioneered by functional programming languages. We briefly describe these below.

\subsubsection{Type, Kernel and Extension Inference}
As described in Section \ref{syntax}, professional end-users tend to prefer languages that do not require explicit type annotations for variables. These languages accomplish this by using dynamic typing, so that types are associated with values instead. Statically-typed functional programming languages have also eliminated type annotations in many cases, however, using a technique known as {\it type inference} \cite{tapl}.

>>>>

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************
\documentclass[10pt, conference, compsocconf]{IEEEtran}

\usepackage{cite}
\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\usepackage{listings}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape,
  basicstyle=\ttfamily\footnotesize,
  morekeywords={lambda, self, assert},
  numbers=left,
  numberstyle=\footnotesize\color{gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
}

\usepackage{float}
\floatstyle{ruled}
\newfloat{codelisting}{tp}{lop}
\floatname{codelisting}{Listing}

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

@@

\usepackage{placeins}

@@

\author{\IEEEauthorblockN{Cyrus Omar, Nathan Fulton and Jonathan Aldrich}
\IEEEauthorblockA{School of Computer Science\\
Carnegie Mellon University\\
Pittsburgh, PA, USA\\
\url{http://www.acelang.org/}
}}

@@

\begin{abstract}
We introduce the Ace compilation environment and demonstrate its suitability as a foundational tool for both research and practice in high-performance computing.
Ace consists of a statically-typed programming language with {\em user-extensible semantics} defined using a compile-time meta\-language, Python. Users specify new primitive types and their operations by equipping type definitions, which are first-class objects in the metalanguage, with  methods that the compiler selectively invokes when type checking and translating expressions, a mechanism we call {\em active type-checking and translation (\ATT)}. We demonstrate the flexibility of this mechanism by implementing primitives from several widely-known languages, including the entirety of OpenCL. Ace also supports more general forms of metaprogramming, and Ace functions can be compiled and launched directly from Python with standard numeric data structures as arguments. Using each of these features, we designed a scientific simulation framework that allows users to modularly specify, compile and orchestrate the execution of parameterized families of scientific simulations on clusters of GPUs. This framework has been used to successfully conduct large-scale, high-performance neuroscience simulations, providing initial evidence that Ace is useful in practice today.
\end{abstract}

\begin{IEEEkeywords}
programming languages, metaprogramming, type systems, extensibility, heterogeneous computing
\end{IEEEkeywords}

@@

Computer-aided simulation and data analysis techniques have transformed science and engineering. Surveys show that scientists and engineers now spend up to 40\% of their time writing software \cite{howison2011scientific, hannay2009scientists}. Most of this software targets desktop hardware and about 20\% of scientists also target either local clusters or super\-computers for more numerically-intensive computations \cite{hannay2009scientists}. To fully harness the power of these platforms, however, these {\em professional end-user developers} \cite{segal2007some} must write parallel programs.

Professional end-users today generally use high-level scripting languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where these interpretted languages are too slow, they will typically call into code written in a statically-typed, low-level language like C or Fortran and use low-level parallel abstractions like pthreads or MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously error-prone and difficult to use, even for expert programmers.

The research community has proposed dozens of novel language features and parallel programming abstractions that aim to  more equitably balance important concerns related to performance, portability, verifiability and ease-of-use. Unfortunately, professional end-users are hesitant to adopt these, with many generally skeptical regarding the practicality of new approaches that originate in the research community. This viewpoint was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding}, who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.'' Although this sentiment is easy to dismiss as paradoxical, we believe that it demands direct examination by researchers working to advance the practice of high-performance computing.

A major barrier to adoption of new language features and abstractions is that they are often developed and distributed together with a new programming language. Some prominent examples of this language-oriented approach include:
\begin{itemize}
\item UPC \cite{upc}, CAF \cite{caf}, Fortress \cite{fortress}, X10 \cite{x10} and Chapel \cite{chapel}, which support different forms of the partitioned global address space (PGAS) model; 
\item Actors, CSPs and other forms of message passing are best supported by languages like Erlang, Charm++, Occam and Axum.
\item Nested data parallelism is best supported by languages like Cilk and NESL.
\end{itemize}

Unfortunately, using a specialized research language comes with a host of practical challenges, such as the cost of learning the language, and the lack of well-developed libraries and tools. In fact, the most widely adopted parallel programming abstractions have been those that are distributed as libraries, such as MapReduce, Global Arrays and MPI. This approach is more practical because it allows developers to progressively introduce parallelism into existing programs, leverage a familiar and well-developed general-purpose programming language when not directly invoking parallel primitives, and utilize different abstractions within a single program without encountering friction inherent to language boundaries. Abstractions developed as libraries are limited, however, in that they cannot alter or augment the semantics of the host language itself. Thus, abstractions that require new type checking rules and compilation logic for the types and associated operations they introduce are at a relative disadvantage in practice.

Ace aims to solve this problem by allowing developers to define and distribute new first-class language constructs in a safely composable manner directly within user libraries using a mechanism called {\em active typechecking and translation (\ATT)}. In particular, \ATT~allows library authors to extend the semantics of the  language while retaining a fixed but flexible grammar. Aided by a form of type inference, the Ace grammar -- a large subset of Python's -- is flexible enough that we are able to define all of the primitives of the C99 and OpenCL languages in a natural manner. These primitives can then be supplemented with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}, as desired. Notably, these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler for a language.

%To enable this and other, Ace programs contain both compile-time and run-time program logic. The compile-time logic, written in the {\em metalanguage}, is the top-level code in each file. Types are first-class objects in the metalanguage, and the logic needed to typecheck and compile the operators associated with a type is bundled with the type definition itself. Abstractions can target existing backend languages, using its existing compilation infrastructure for low-level optimizations, so only the high-level semantics need to be encoded by researchers. Consistent with its design intent, we demonstrate the use of OpenCL as a flexible backend language and compiler framework \cite{opencl}.
%Here, users can define and import new primitive types and operators, as well as do other types of metaprogramming tasks like generate functions programmatically, perform arbitrary code transformations.
% Ace also permits an interactive mode of use whereby the functions can also be executed immediately after being defined, a common usage scenario for professional end-users.

Although the \ATT~mechanism, a key feature of Ace, is designed for use by researchers and domain experts, Ace is designed to be highly practical for professional end-user developers as well. It is based on the Python programming language, which enjoys broad adoption in this community. Indeed, the language itself is implemented entirely as a Python library, so many existing tools, such as editor modes, documentation tools and style checkers, are available immediately.
Ace programs can be generated programmatically and called directly from Python, with numeric arrays as arguments, so Ace integrates immediately with existing workflows to speed up performance-critical portions of code that may benefit from static compilation and support for parallelism.
We demonstrate several of these features by developing a scientific simulation framework, which we use for specifying and orchestrating families of large-scale neurobiological circuit simulations on clusters of GPUs, in Section \ref{clegans}.

@@

%We then introduce a new programming language named Ace (pronounced ``C eloquence'') in Section 3. Ace initially targets code paths where low-level languages, particularly OpenCL, would be used today. By employing a collection of powerful front-end language and compiler design techniques, making pragmatic design choices, and building on top of established infrastructure, we argue that Ace may uniquely satisfy many of the criteria established in Section 2, particularly those related to ease of use, familiarity, performance and tool support.
%
%Unlike many existing languages, Ace is designed to be highly extensible from inception, employing a powerful compile-time language extension and code generation mechanism based around the concept of {\it active libraries} \cite{activelibraries}. We argue that this feature may make Ace particularly suitable as a platform for researchers developing new parallel abstractions and other specialized, high-level language features, as well as  development tools.
%
%In Section 4 we briefly describe a case study where the language was used to conduct large-scale spiking neural circuit simulations on GPU hardware, demonstrating the feasibility of this approach on a non-trivial problem. Finally, conclude in Section 6 with future directions for both the Ace language and the research community.
%

\begin{codelisting}
\begin{lstlisting}
from ace.OpenCL import OpenCL, printf

print "Hello, compile-time world!"

@OpenCL.fn
def main():
    printf("Hello, run-time world!")
main = main.compile()

print "Goodbye, compile-time world!"
\end{lstlisting}
\caption{\texttt{hello.py}}
\label{hello}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acecc hello.py`
Hello, compile-time world.
Goodbye, compile-time world.
$ `cat hello.cl`
__kernel void main() {
    printf("Hello, run-time world.");
}\end{lstlisting}
\caption{Compiling \texttt{hello.py} using \texttt{acecc}.}
\label{helloout}
\end{codelisting}

\section{Ace for Professional End-Users}
%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly. The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}. The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar. Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}. Ace programs are statically-typed, and types themselves are objects in the metalanguage. They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled. Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
\subsection{Example 1: Hello, World!}
To demonstrate the structure of programs and libraries in Ace, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the OpenCL module distributed with Ace. We note that the OpenCL module, which we use throughout this paper in examples, is simply a library like any other. The core of Ace gives no special treatment to it; it is distributed with Ace for convenience.

Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain \verb|print| statements that are executed at {\em compile-time}. Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages. In other words, Python is the {\em compile-time metalanguage} of Ace. In other words, Ace programs are embedded within Python. By making this choice, Ace can  use Python's well-developed module system and associated distribution infrastructure directly (Line 1).

The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive imported from the \verb|ace.OpenCL| module on Line 1. The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a {\em actively-typed} Ace function containing run-time logic written for the \verb|OpenCL| backend, also imported on Line 1. Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)

On Line 8, the {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} which we also call \verb|main| (overwriting the previous definition). The distinction between generic and concrete functions will be explained in our next example. For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acecc|, shown operating at the shell in Listing \ref{helloout}. The \verb|acecc| compiler operates as follows:
\begin{enumerate}
\item Executes the provided Python module (\verb|hello.py|)
\item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment. Each backend may produce one or more files (here, only \verb|hello.cl| is produced).
\end{enumerate}

Use of the \verb|acecc| executable is simply a convenience; the generated code is also available as an attribute of the concrete function: \verb|main.code|. We will also show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, functions can be executed directly, without an intermediate compilation step, if desired.

\subsection{Example 2: Higher-Order Map for OpenCL}
\begin{codelisting}
\begin{lstlisting}
from ace.OpenCL import OpenCL, get_global_id

@OpenCL.fn
def map(input, output, f):
    gid = get_global_id(0)
    output[gid] = f(input[gid])
\end{lstlisting}
\caption{[\texttt{listing3.py}] A generic data-parallel higher-order map function written using the user-defined Ace OpenCL module.}
\label{map}
\end{codelisting}

\begin{codelisting}
\begin{lstlisting}
from listing3 import map
from ace.OpenCL import global_ptr, double, int

@OpenCL.fn
def add5(x):
    return x + 5

A, B = global_ptr(double), global_ptr(int)
map_add5_dbl = map.compile(A, A, add5.ace_type)
map_add5_int = map.compile(B, B, add5.ace_type)
\end{lstlisting}
\caption{[\texttt{listing4.py}] The generic map function compiled to map the sin function over global arrays of double-precision floating-point numbers. The \texttt{get\_global\_id} and \texttt{sin} functions are OpenCL primitives.}
\label{mapadd5dbl}
\end{codelisting}

\begin{codelisting}
\begin{lstlisting}[style=OpenCL]
#pragma OPENCL EXTENSION cl_khr_fp64 : enable

double add5(double x) {
    return x + 5;
}

__kernel void map_add5_dbl(
  __global double* input, 
  __global double* output) 
{
    size_t gid;
    gid = get_global_id(0);
    output[gid] = add5(input[gid]);
}

int add5__1(int x) {
    return x + 5;
}
